### 2011-12-20 paper_draft.tex
reworking of the draft, more references

\subsection*{openSNP}
We created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints. Users are encouraged to list as many phenotypes as possible through to a simple achievement-system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile page.

The possible answers in terms of variations for a single phenotype are not limited and every user can add completely new phenotypes if the corresponding questions about this are lacking. To reduce the amount of manual data curation openSNP tries to avoid the entry of the same phenotype or variation, but with a slightly different spelling by helping users at entering data by an autocompletion-feature which lists similar entries which are already in the openSNP-database.

On the side of getting access to the data users can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation or can access a single download that includes all genotyping-files and all phenotypes in a comma separated table. Additionally users can access the genetic data through the Distributed Annotation System, which allows to get all data for specific chromosomes and specific positions on single chromosomes. Users can discuss SNPs and phenotypes on the platform using a simple commenting-system or private messages.

Between the start of openSNP on 09/27/2011 and 12/18/2011 214 people have signed up with openSNP, 79 of those have uploaded their genotyping files. Through this the openSNP database lists 69486471 SNPs which are distributed over 1938604 unique Rs-IDs. In the same timeframe all users combined have entered 675 variations which are distributed over 47 different phenotypes. See figure n for a distribution of data acquisition over time.

The mean number of users that have entered their variation for a single phenotype is 14.36 (SD 12.65), the median is 10. The distribution of how many users have entered their data per phenotype can be seen in figure n+1. The phenotype which has been provided by the most users is the eye color which has been provided by 54 different users. There are two phenotypes which have only been provided by a single user: The score of the Writing-SAT and triglyceride-levels.

A total number of 15229 documents relevant to the SNP-IDs which are listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. Of the primary literature 25 \% are released in Open Access-journals and can be freely free of charge by every user (Figure n+2). For usability reasons SNPs are ranked by the amount of information gathered by the external services.

The external services themselves are ranked by how easy users can access information out of these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science. Lowest values are given to the Mendeley-results, as those aren't necessary freely available for every user. A entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley-entry.

\section*{Discussion}
Although prices for exome or even full genome sequencing are dropping rapidly, in comparison to this GWAS tend to stay cheaper and offer a possibility to get insights on genetic variations on a population level and allow an analysis of SNPs which are linked to different traits. Despite this benefit in terms of costs it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow the detection of the causes for this and needs a large enough sample which has to be statistically analysed by using sound methods. Nevertheless, GWAS are still frequently used and new associations still can be found (we should list some recent results here -> oxytocin maybe? we could look at the twitter-timeline of @opensnporg, i publish papers i find there).

One way to bring costs for GWAS even further down is to make use of already available genotyping results and datasets. Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for those genotypings by themselves. As we have found in the survey on the sharing such results many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.

With openSNP we've build a platform that can be used by customers of DTC genetic testing to share their data to easily share their genetic and phenotypic data with a wide audience as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. Customers of DTC genetic testing also benefit of an easy access to primary literature on SNPs and genetic variation they carry. While we don't have collected enough data to perform full scale GWAS yet this might be possible in the future, as user numbers are rising.

By crowdsourcing the acquisition of genetic and phenotypic data openSNP faces the same problems as any other open platform on the Internet, namely we have to trust users about the data they upload and enter on openSNP. Additionally the quality of the data varies, especially in terms of accuracy on the phenotypic variation. While we try to suggest similar entries to users there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or varations in traits may arise.

There are two possible solutions to this problem: One could only allow some (trusted) users to enter new phenotypes or one could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add variations that have not been available from the start on. Both methods have a disadvantage: In either case it makes it harder for users to enter their data and by this highers the bar for participation, which ultimately could lead to less data entered. Facing this trade-off we decided to keep entering data easy, at the cost that users who want to perform GWAS with the data need to perform more quality control.

Another thing that should kept in mind is a possible bias in which data is available on openSNP, as we can't rule out the possibility that only a biased subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.

The advent of DTC genetic testing has lead to new ethical and social issues. Much of the critique on DTC genetic testing focusses on the practice of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information and help them to put the new knowledge to good use. Less attention has been given to the privacy implications that come with this cheap way of obtaining genetical information.

By making the own genetic and medical information public it becomes clear that the issue of privacy needs to get a closer view. Our survey has shown that people are concerned about their privacy and are concerned that stakeholders like employers, insurance companies, governments or advertisers misuse this information. Policy makers start to react to those changes by introducing laws like the \textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact widely available genetic information. DTC genetic testing companies on their side also try to educate their customers about the risks of releasing genetic data.

We transparently address the problem of privacy implications that come with releasing the data twice, during registering for openSNP and during the upload of the DTC results. Users have to check that they have read and understood the disclaimer about possible side-effects that can arise by making their data public. To further improve this process and get informed consent we are working to implement the procedures which are currently developed by \textit{Consent for Research} (http://www.weconsent.us).

@@

\subsection*{Technical realization of the platform}
The main platform is realized using the web-framework Ruby on Rails. Postgres is used as the main database backend for Rails. The database stores genotyping results, phenotypic information of the users, literature results of Mendeley, the Public Library of Science and summaries on SNPs which can be found in SNPedia. The literature database of Mendeley is queried using the REST API, which delivers results in JSON. The literature database of the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format. Summaries on SNPs are provided by SNPedia, through querying the content through the MediaWiki-API. All databases are queried using the unique identifier of each SNP as search term.

SNPs are cataloged by their unique identifier, which consists of a prefix (mostly \textit{rs}) and a unique number. This is a common format which is used by the NCBI dbSNP database and is also widely used and easily parsed from different literature-sources. Publications out of the different databases and the genotypes of the users are associated with single SNPs by the Rs-ID. Allele and genotype frequencies are are updated regularly, based on the data present in openSNP.

Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are be mailed to users and queries to external resources are handled using the ruby-gem Resque and a Redis-server. Search-features on the platform itself are implemented using SOLR and the ruby-gem sunspot. Additionally data can be requested from openSNP using the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}. The data for this is stored in a mySQL-database, the delivery of the data is done by using ProServer \cite{Finn2007}, which was modified by Gel et al. for use in easyDAS \cite{GelMoreno2011}.

A flowchart of all services incorporated in openSNP and how users can upload or access the data is given in figure n+3. The source code of openSNP is published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

>>>>

\subsection*{openSNP}
We created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints. Users are encouraged to list as many phenotypes as possible through a simple achievement system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile pages.

The specification of variations on a single phenotype and the phenotypes themselves is open and not limited to pre-defined categories. Every user can add completely new phenotypes that are not yet queried by openSNP. To reduce the amount of manual data curation, openSNP tries to harmonize the expression and spelling of the same phenotype or variation by implementing an autocompletion-feature which lists similar entries to what is being entered that are already in the openSNP database.

OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypes in a comma-separated table. Additionally, the genetic data is accessible through the Distributed Annotation System, which offers all data for specific chromosomes and specific positions on single chromosomes. Users can discuss SNPs and phenotypes on the platform using a simple commenting system or private messages.

Between the start of openSNP on 09/27/2011 and 12/18/2011, 214 people have signed up with openSNP, 79 of which have uploaded their genotyping files. The openSNP database thereby lists 69,486,471 SNPs, which are distributed over 1,938,604 unique Rs-IDs. In the same timeframe all users combined have entered a total of 675 variations on 47 different phenotypes. See figure n for a distribution of data acquisition over time.

The mean number of users that have entered their variations for a single phenotype is 14.36 (SD 12.65), the median is 10. The distribution of how many users have entered their data per phenotype can be seen in figure n+1. The phenotype provided by the most users is the eye color, which has been entered by 54 users. There are two phenotypes which have so far only been provided by a single user: The SAT Writing score and triglyceride-levels.

In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure n+2). For usability reasons, SNPs are ranked by the amount of information gathered by the external services.

The external services themselves are ranked by how easily users can access information from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science. Lowest values are given to the Mendeley results, as those aren't necessary freely available to every user. An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.

\section*{Discussion}
Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper and offer a possibility of getting insight into genetic variations on a population level and allow an analysis of SNPs which are linked to different traits. Despite this benefit in terms of cost, it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow inference on the cause for any correlation. Furthermore, GWAS need a large enough sample which has to be statistically analysed by using sound methods. Nevertheless, GWAS are still frequently used and new associations still can be found (we should list some recent results here -> oxytocin maybe? we could look at the twitter-timeline of @opensnporg, i publish papers i find there).

One way of bringing down costs for GWAS even further is to make use of already available genotyping results and datasets. Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for the genotyping by themselves. As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.

With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.

By crowdsourcing the acquisition of genetic and phenotypic data, openSNP faces the same problems as any other open platform on the Internet, namely the need to trust users regarding the data they upload and enter on openSNP. Additionally, the quality of the data varies, especially in terms of accuracy on the phenotypic variation. While we try to suggest similar entries to the users, there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or varations in traits may arise.

There are two possible solutions to this problem: One could only allow some (trusted) users to enter new phenotypes or one could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add variations that have not been available from the start. Both methods have a disadvantage: In either case it makes it harder for users to enter their data and thereby raises the bar for participation, which ultimately could lead to less data entered. Facing this trade-off, we decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform more quality control.

Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.

The advent of DTC genetic testing has led to new ethical and social issues. Much of the critique on DTC genetic testing focusses on the practice of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}. Less attention has been paid to the privacy implications that come with this cheap way of obtaining genetical information \cite{Caulfield2011,Joh2011}.

With a move towards making one's own genetic and medical information public, it becomes clear that the issue of privacy needs to get a closer look. Our survey has shown that people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments or advertisers might misuse the information. Policy makers start to react to those changes by introducing laws like the \textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.

We openly address the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects that can arise by publishing their data. To further improve this process and get informed consent, we are working on implementing the procedures which are currently developed by \textit{Consent for Research} (http://www.weconsent.us).

@@

\subsection*{Technical realization of the platform}
The main platform is realized using the web framework Ruby on Rails. Postgres is used as the main database backend for Rails. The database stores genotyping results, users' phenotypic information, literature results from Mendeley and the Public Library of Science as well as summaries on SNPs which can be found in SNPedia. The literature database of Mendeley is queried using the REST API, which delivers results in JSON. The literature database of the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format. Summaries on SNPs are provided by SNPedia, through querying the content via the MediaWiki API. All databases are queried using the unique identifier of each SNP as the search term.

SNPs are cataloged by their unique identifier, which consists of a prefix (mostly \textit{rs}) and a unique number. This is a common format, which is employed by the NCBI dbSNP database and is also widely used and easily parsed from different literature sources. Publications from the different databases as well as the users' genotypes are associated with individual SNPs by the Rs-ID. Allele and genotype frequencies are updated regularly, based on the data present in openSNP.

Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are to be mailed to users and queries to external resources are handled using the ruby gem Resque and a Redis server. Search features on the platform itself are implemented using SOLR and the ruby gem Sunspot. Additionally, data can be requested from openSNP using the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}. The required data is stored in a mySQL database, the delivery of the data is done by using ProServer \cite{Finn2007}, which was modified by Gel et al. for use in easyDAS \cite{GelMoreno2011}.

A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure n+3. The source code of openSNP is published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
